Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-11 manuscript/manuscript.Rmd
first cut of methods

##Data

##Conditional Probability Analysis
We used a conditional probability analysis approach to explore associations between chlorophyll *a* concentrations  and World Health Organization (WHO) and U.S. Environmental Protection Agency (U.S. EPA) microcystin-LR health advisory levels [REFS].  Many levels have been suggested [Table \ref{tab:microcystin_levels}], but lakes with higher microcystin-LR concentrations in the NLA were rare. Only `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat)),2)` % of lakes sampled had a concentration greater than 10.  Thus, for this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were  0.3, 1, 1.6, and 2 $\mu$g/L.

@@

```{r microcystin_levels,echo=FALSE,results='asis', fig.cap="Various suggested microcystin-LR health advisory concentrations. \\label{fig:microcystin_levels}"}
source_of_stnd <- c("WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "U.S. EPA",
                    "U.S. EPA")
type_of_stnd <- c("Recreational",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Drinking",
                  "Drinking",
                  "Drinking")
concentration <- c("2-4 $\\mu$g/L",
                   "10-20 $\\mu$g/L",
                   "20-2000 $\\mu$g/L",
                   ">2000 $\\mu$g/L",
                   "1 $\\mu$g/L",
                   "0.3 $\\mu$g/L",
                   "1.6 $\\mu$g/L")
#note <- c("Low probability of adverse health effects",
#          "Moderate probability of adverse health effects",
#          "High probability of adverse health effects, These higher #concentrations typically only seen in visible scums/algal mats",
#          "Very high probability of adverse health effects",
#          "",
#          "Children younger than school age",
#          "Individuals School age and older")
mc_conc_table <- data.frame(source_of_stnd, type_of_stnd, concentration) %>%
  arrange(type_of_stnd)
names(mc_conc_table)<-c("Source","Type","Concentration")
pandoc.table(mc_conc_table,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

@@

```{r mc_chla_table, results='asis', echo=FALSE, fig.cap="Chlorophyll \textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{fig:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

>>>>

##Data
We used the 2007 National Lakes Assessment (NLA) water quality and microcystin-LR concentration data [REF].  The NLA is a probability based design that, when using the provided sample weights, can be used to provide unbiased estimates of lake condition for all lakes in the original sample frame (e.g. lakes greater than 4ha in the conterminous US).  These data represent a snapshot of water quality from the summer of 2007 and data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.  All results are reported using weighted estimates.

##Conditional Probability Analysis
We used a conditional probability analysis (CPA) approach to explore associations between chlorophyll *a* concentrations  and World Health Organization (WHO) and U.S. Environmental Protection Agency (U.S. EPA) microcystin-LR health advisory levels [REFS].  Many levels have been suggested [Table \ref{tab:microcystin_levels}], but lakes with higher microcystin-LR concentrations in the NLA were rare. Only `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat)),2)` % of lakes sampled had a concentration greater than 10.  Thus, for this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were  0.3, 1, 1.6, and 2 $\mu$g/L.

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-REF] and Hollister et al. [-REF] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with the upper confidence level being 50% or greater.  Details of the specific implementation we used are available at https://github.com/USEPA/microcystinchla.

@@

```{r microcystin_levels,echo=FALSE,results='asis', fig.cap="Various suggested microcystin-LR health advisory concentrations. \\label{tab:microcystin_levels}"}
source_of_stnd <- c("WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "WHO",
                    "U.S. EPA",
                    "U.S. EPA")
type_of_stnd <- c("Recreational",
                  "Recreational",
                  "Recreational",
                  "Recreational",
                  "Drinking",
                  "Drinking",
                  "Drinking")
concentration <- c("2-4 $\\mu$g/L",
                   "10-20 $\\mu$g/L",
                   "20-2000 $\\mu$g/L",
                   ">2000 $\\mu$g/L",
                   "1 $\\mu$g/L",
                   "0.3 $\\mu$g/L",
                   "1.6 $\\mu$g/L")
#note <- c("Low probability of adverse health effects",
#          "Moderate probability of adverse health effects",
#          "High probability of adverse health effects, These higher #concentrations typically only seen in visible scums/algal mats",
#          "Very high probability of adverse health effects",
#          "",
#          "Children younger than school age",
#          "Individuals School age and older")
mc_conc_table <- data.frame(source_of_stnd, type_of_stnd, concentration) %>%
  arrange(type_of_stnd)
names(mc_conc_table)<-c("Source","Type","Concentration")
pandoc.table(mc_conc_table,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

@@

```{r mc_chla_table, results='asis', echo=FALSE, fig.cap="Chlorophyll \textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
```

### 2015-09-11 manuscript/manuscript.tex
first cut of methods

@@

We used a conditional probability analysis approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and U.S. Environmental Protection Agency (U.S.
EPA) microcystin-LR health advisory levels {[}REFS{]}. Many levels have
been suggested {[}Table \ref{tab:microcystin_levels}{]}, but lakes with
higher microcystin-LR concentrations in the NLA were rare. Only 1.16 \%
of lakes sampled had a concentration greater than 10. Thus, for this
analysis we focus on the microcystin concentrations that are better
represented in the NLA data. These were 0.3, 1, 1.6, and 2 \(\mu\)g/L.

@@

\begin{verbatim}
## Loading required package: grid
\end{verbatim}

>>>>

We used the 2007 National Lakes Assessment (NLA) water quality and
microcystin-LR concentration data {[}REF{]}. The NLA is a probability
based design that, when using the provided sample weights, can be used
to provide unbiased estimates of lake condition for all lakes in the
original sample frame (e.g.~lakes greater than 4ha in the conterminous
US). These data represent a snapshot of water quality from the summer of
2007 and data on chlorophyll \emph{a} and microcystin-LR concentrations
are available for lakes. All results are reported using weighted
estimates.

@@

We used a conditional probability analysis (CPA) approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and U.S. Environmental Protection Agency (U.S.
EPA) microcystin-LR health advisory levels {[}REFS{]}. Many levels have
been suggested {[}Table \ref{tab:microcystin_levels}{]}, but lakes with
higher microcystin-LR concentrations in the NLA were rare. Only 1.16 \%
of lakes sampled had a concentration greater than 10. Thus, for this
analysis we focus on the microcystin concentrations that are better
represented in the NLA data. These were 0.3, 1, 1.6, and 2 \(\mu\)g/L.

A detailed discussion of CPA is beyond the scope of this paper, but see
Paul et al. {[}-REF{]} and Hollister et al. {[}-REF{]} for details. For
this analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with the upper
confidence level being 50\% or greater. Details of the specific
implementation we used are available at
\url{https://github.com/USEPA/microcystinchla}.

@@
